Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-12-10 inst/manuscript/manuscript.tex
some upates

\Abstract{
Non-Gaussian data are common in practice and include binary, count,
skewed, and proportion data types. Often, non-Gaussian data are modeled
using a generalized linear model (GLM). GLMs typically assume that
observations are independent of one another. This is an impractical
assumption for spatial data, as nearby observations tend to be more
similar than distant ones. The \pkg{spmodel} package in \proglang{R}
provides a suite of tools for fitting spatial generalized linear models
(SPGLMs) to non-Gaussian data and making spatial predictions (i.e.,
Kriging). SPGLMs for point-referenced (x- and y-coordinates) support are
fit using the \code{spglm()} function, while SPGLMs for areal (lattice,
polygon) support are fit using the \code{spgautor()} function. Both
\code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood
which marginalizes over the model's fixed effects and latent mean while
formally incorporating spatial covariance among observations. The inputs
and outputs of \code{spglm()} and \code{spgautor()} closely resemble the
\code{glm()} function from base \proglang{R}, easing the transition from
GLMs to SPGLMs. \pkg{spmodel} provides and builds upon several commonly
used helper functions for model building like \code{summary()},
\code{plot()}, and \code{fitted()}, among others. Spatial predictions of
the latent mean at unobserved locations are obtained using
\code{predict()} or \code{augment()}. \pkg{spmodel} accommodates myriad
advanced modeling features like geometric anisotrpoy, nonspatial random
effects, analysis of variance, and more. Throughout, we use
\pkg{spmodel} to fit SPGLMs to moose presence and counts in Alaska,
United States (US), skewed conductivity data in the Southwestern US,
harbor seal abundance trends in Alaska, US, and voter turnout rates in
Texas, US.
}

@@

\Address{
    Michael Dumelle\\
    United States\\
Environmental Protection Agency\\
    200 SW 35th St\\
Corvallis, OR, 97330\\
  E-mail: \email{Dumelle.Michael@epa.gov}\\

@@

\linenumbers

@@

The standard GLM assumes the elements of \(\mathbf{y}\) are independent.
This independence assumption is typically impractical for spatial data.
For spatial data, nearby observations tend to be more similar than
distant observations \citep{tobler1970computer}, which leads to positive
spatial covariance among observations. The consequences of ignoring
spatial covariance in statistical models for spatial data can be severe
and include imprecise parameter estimates as well as misleading standard
errors that inflate Type-I error rates and decrease power
\citep{zimmerman2024spatial}.

@@

Through inclusion of \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\), the spatial GLM (SPGLM) can be written as
\begin{align}\label{eq-spglm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
\end{align} Assuming independence among \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\), it follows that \begin{align}\label{eq-spcov}
 \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
\end{align} Henceforth, we refer to \(\sigma^2_{\tau}\) as
\(\sigma^2_{de}\) (for spatially dependent error variance) and
\(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for independent error
variance). The parameters \(\sigma^2_{de}\), \(\sigma^2_{ie}\), and
\(\phi\), in addition to any other parameters in \(\mathbf{R}\), compose
\(\boldsymbol{\theta}\), the covariance parameter vector.

@@

The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
recently released a full set of modeling tools for SPGLMs fit using the
novel Laplace approach described by \citet{ver2024marginal}. These
modeling tools are approachable and mirror the familiar \code{glm()}
syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs
relatively seamless. The \code{spglm()} function fits SPGLMs for
point-referenced support (e.g., x- and y-coordinates representing point
locations in a field; these models are sometimes called
``geostatistical'' models), while the \code{spgautor()} function fits
SPGLMs for areal support (e.g., polygon boundaries representing
geographic subsets of a region; these models are sometimes called
``autoregressive'' models). For both point-referenced and areal
supports, \pkg{spmodel} supports the binomial distribution for binary
data, Poisson and negative binomial distributions for count data, Gamma
and inverse Gaussian distributions for skewed data, and the beta
distribution for proportion data. There are 20 different spatial
covariance structures available including the exponential, Gaussian, and
spherical for point-referenced support (Figure\(~\)\ref{fig-type}) and
the conditional autoregressive, and simultaneous autoregressive
structures for areal support \pkg{spmodel} provides tools for commonly
used model summaries, visualizations, and diagnostics (e.g., fitted
values) using standard \proglang{R} helper functions like
\code{summary()}, \code{plot()}, and \code{fitted()}, among others.
\pkg{spmodel} also provides tools to predict \(\mathbf{w}\) at new
locations and quantify uncertainty in those prediction using
\code{predict()} and \code{augment()}. This core functionality, combined
with several advanced features we describe throughout the manuscript,
enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools
previously missing from the existing \proglang{R} ecosystem.

@@

Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version
0.11.1) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in
terms of scope and feel. Both packages use similar syntax as
\code{glm()}, accommodate flexible \code{formula} arguments (e.g.,
offsets, splines), handle spatial covariance that decays at different
rates in different directions (i.e., geometric anisotropy), incorporate
nonspatial random effects, support other \proglang{R} packages for
modeling like \pkg{broom} \citep{robinson2021broom, kuhn2022tidy},
\pkg{emmeans} \citep{lenth2024emmeans}, and \pkg{car}
\citep{fox2019car}, and have tools for model summaries, prediction, and
simulating data. There are some notable differences between the two
packages, however. \pkg{sdmTMB} supports several additional GLM
distributions like the Tweedie, supports Hurdle models, and can
incorporate prior information through Bayesian applications.
\pkg{sdmTMB} also provides tools for working with temporal data and
spatiotemporal data and provides enhanced visualizations of the model's
marginal effects. \pkg{sdmTMB} does require a preprocessing step of
constructing a mesh prior to model fitting (using the stochastic partial
differential equation approach), and the density of the mesh can affect
model results and computational complexity. On the other hand,
\pkg{spmodel} does not require the construction of a mesh prior to model
fitting. \pkg{spmodel} supports 20 different spatial covariances and
models them directly, rather than using a precision matrix approximation
to the Matérn spatial covariance as in \pkg{sdmTMB}. \pkg{spmodel} can
model data directly using neighborhood distance and autoregressive
models, rather than relying on the polygon centroid (as in
\pkg{sdmTMB}), which may not be within the polygon's boundaries.
\pkg{spmodel} provides experimental design tools (e.g., analysis of
variance, contrasts), supports \pkg{sf} objects in modeling and
prediction functions \citep{pebesma2018sf}, has several specialized
model diagnostics like leverage values and Cook's distances, and has
analytic solutions for fixed effect and prediction standard errors.
Other similarities and differences do exist between \pkg{sdmTMB} and
\pkg{spmodel}, and both packages continue to evolve. Overall, we believe
that these packages are complementary and enhance the suite of SPGLM
tools accessible to practitioners.

@@

We can estimate the fixed effects using generalized least squares (GLS)
principles, a common practice for linear models estimated using
restricted maximum likelihood methods. Had we observed \(\mathbf{w}\), a
GLS estimator for \(\boldsymbol{\beta}\) is given by
\begin{align}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{align} where
\(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
While we do not \(\mathbf{w}\), we do estimate it via
\(\hat{\mathbf{w}}\), and thus it is reasonable to define
\(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). To derive
properties of \(\hat{\boldsymbol{\beta}}\) like expectation and
variance, we must derive these properties for \(\hat{\mathbf{w}}\) by
conditioning on \(\mathbf{w}\) as if it were observed and leveraging the
laws of total expectation and variance. Based on asymptotic properties
of (restricted) maximum likelihood estimators
\citep{cressie1993asymptotic}, we may assume that given \(\mathbf{w}\),
\(\hat{\mathbf{w}}\) has mean \(\mathbf{w}\) and variance approximately
equal to \(-\mathbf{H}^{-1}\), the negative inverse Hessian (i.e., the
inverse observed information matrix). Thus it follows that
\(\text{E}(\hat{\mathbf{w}})\) is given by \begin{align}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{align} and \(\text{Var}(\hat{\mathbf{w}})\) is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & \approx \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align} Putting this all together, it follows that
\(\hat{\boldsymbol{\beta}}\) is unbiased for \(\boldsymbol{\beta}\):
\begin{align}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
\end{align} Moreover, it follows that \begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
\end{align} In practice, \(\text{Var}(\hat{\boldsymbol{\beta}})\) is
estimated by evaluating \(\boldsymbol{\Sigma}\) at
\(\hat{\boldsymbol{\theta}}\), the estimated covariance parameter
vector.

@@

Inspecting model diagnostics is an important step of the modeling
process that can yield valuable insights into model behavior and unusual
observations. \citet{montgomery2021introduction} contextualize three
components of unusual observations: outliers, leverage, and influence.
An observation is an outlier if it has an extreme response value
relative to expectation. The response GLM residuals simply compare the
observation to its fitted latent mean: \begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align} Because observations often have a unique support in a GLM
(e.g., only two possible response values for binary data) and the
variance of an observation generally depends on its mean, response
residuals lack some utility. Deviance residuals are a function of
response residuals that are appropriately scaled to behave more like
response residuals in a standard linear model. Deviance residuals are
given by \begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align} where \(\mathbf{d}\) is a vector of individual deviances.
The sum of the squared deviance residuals equals the sum of the elements
of \(\mathbf{d}\), known as the deviance of the model fit. The deviance
of the model fit quantifies twice the difference in log likelihoods
between the a saturated model that fits every observation perfectly
(i.e., \(\text{y}_i = f^{-1}(\hat{\text{w}}_i)\) for all \(i\)) and the
fitted model \citep{myers2012generalized}. Deviance is often used as a
fit statistic; lower values of deviance imply a better model fit
(compared to the observed data). Pearson and standardized residuals are
other types of GLM residuals that involve a scaling of the response
residuals. The Pearson residuals scale \(\mathbf{r}_{r}\) by the square
root of \(\mathbf{V}\), a diagonal matrix with \(i\)th diagonal element
equal to the variance of the response distribution evaluated at
\(f^{-1}(\hat{\text{w}}_i)\) \citep{faraway2016extending};
\(\mathbf{V}\) is sometimes called the GLM weight matrix. The
standardized residuals scale the deviance residuals by
\(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where \(\mathbf{L}_{ii}\) is
the \(i\)th diagonal element of the leverage matrix, which we discuss
next.

An observation has high leverage if its combination of explanatory
variables is far away from other observations. In a linear model, the
leverage (i.e., hat) values are the diagonal of the leverage (i.e.,
projection, hat) matrix,
\(\mathbf{L} = \mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top\).
In a GLM, the leverage matrix is given by \begin{align}
  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2}.
\end{align} The larger the value of \(\mathbf{L}_{ii}\), the more severe
the leverage from the \(i\)th observation.

@@

The \code{moose} data in \pkg{spmodel} contain information on moose
(Alces Alces) presence in the Togiak region of Alaska, USA. \code{moose}
is an \code{sf} object, a special data frame that is supplemented with
spatial information using the \pkg{sf} package in \proglang{R}
\citep{pebesma2018sf}. After loading \pkg{spmodel}, the first few rows
of \code{moose} look like:

@@

Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel}
helper functions are available and that each of the \code{spglm()}
models uses the same hierarchical likelihood
(Equation\textasciitilde{}\ref{eq-marginal}):

@@

The \code{plot()} function can also be used to return similar
diagnostics as from \code{lm()} and \code{glm()}, with additional tools
for diagnosing spatial covariance. For example, we can inspect Cook's
distance values and the empirical spatial covariance as a function of
distance with (Figure\(~\)\ref{fig-sp-diagnostic2}):

@@

In addition to moose presence, moose counts are also recorded in
\code{moose} (Figure\(~\)\ref{fig-moose-data-count}). The Poisson and
negative binomial response distributions can be used to model SPGLMs for
count data. The Poisson distribution mean is equal to its variance,
while the negative binomial has an extra parameter to accommodate
overdispersion (where the variance is larger than the mean). Using a
spherical spatial covariance function, we may fit both a Poisson and
negative binomial SPGLM changing the \code{family} argument:

@@

To model spatial dependence, we used a conditional autoregressive
function. Conditional and simultaneous autoregressive functions
characterize spatial distance through neighborhood relationships (rather
than Euclidean distance) and have \code{spcov_type} values of
\code{"car"} and \code{"sar"}, respectively. By default, Queen's
distance is used to determine whether two sites are neighbors, though
custom neighborhood matrices can be passed via \code{W}. Row
standardization is also assumed by default; this can be changed via
\code{row_st}. Using \code{random}, we also specified a nonspatial
random effect for seal stock, which implies seals belonging to the same
stock share extra covariance. The \code{random} argument uses similar
syntax as \pkg{lme4} \citep{bates2015lme4} and \pkg{nlme}
\citep{pinheiro2006mixed} to specify nonspatial random effects.

@@

According to AIC, the geostatistical SPGLM:

@@

The default estimation method in \pkg{spmodel} for SPGLMs is restricted
maximum likelihood (REML), while maximum likelihood (ML) can also be
used. A benefit of benefit of REML is that it can yield unbiased
estimates of covariance parameters \citep{cressie1993asymptotic}. A
drawback of REML is that likelihood-based statistics are often
considered invalid when the models have different explanatory variable
or fixed effect structures \citep{wolfinger1993covariance}, though
\citet{gurka2006selecting} provides some evidence to the contrary. In
contrast to REML estimators, ML estimators are generally biased for
covariance parameters, though in practice this bias tends to be small.
Moreover, when using ML, likelihood-based comparisons are valid for
models with different explanatory variable or fixed effect structures.
Using ML, we can evaluate the significance of log income on voter
turnout using a likelihood ratio test (for nested models):

\begin{CodeChunk}
\begin{CodeInput}
R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
R> spbeta_reduced_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
R> anova(spbeta_full_ml, spbeta_reduced_ml)
\end{CodeInput}
\begin{CodeOutput}
Likelihood Ratio Test

>>>>

\Abstract{
Non-Gaussian data are common in practice and include binary, count,
skewed, and proportion structures. Often, non-Gaussian data are modeled
using a generalized linear model (GLM). GLMs typically assume that
observations are independent of one another. This is an impractical
assumption for spatial data, as nearby observations tend to be more
similar than distant ones. The \pkg{spmodel} package in \proglang{R}
provides a suite of tools for fitting spatial generalized linear models
(SPGLMs) to non-Gaussian data and making spatial predictions (i.e.,
Kriging). SPGLMs for point-referenced (x- and y-coordinates) support are
fit using the \code{spglm()} function, while SPGLMs for areal (lattice,
polygon) support are fit using the \code{spgautor()} function. Both
\code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood
which marginalizes over the model's fixed effects and latent mean while
formally incorporating spatial covariance among observations. The inputs
and outputs of \code{spglm()} and \code{spgautor()} closely resemble the
\code{glm()} function from base \proglang{R}, easing the transition from
GLMs to SPGLMs. \pkg{spmodel} provides and builds upon several commonly
used helper functions for model building like \code{summary()},
\code{plot()}, and \code{fitted()}, among others. Spatial predictions of
the latent mean at unobserved locations are obtained using
\code{predict()} or \code{augment()}. \pkg{spmodel} accommodates myriad
advanced modeling features like geometric anisotrpoy, nonspatial random
effects, analysis of variance, and more. Throughout, we use
\pkg{spmodel} to fit SPGLMs to moose presence and counts in Alaska,
United States (US), skewed conductivity data in the Southwestern US,
harbor seal abundance trends in Alaska, US, and voter turnout rates in
Texas, US.
}

@@

\Address{
    Michael Dumelle\\
    United States\\
Environmental Protection Agency\\
    Office of Research and Development\\
200 SW 35th St\\
Corvallis, OR, 97330\\
  E-mail: \email{Dumelle.Michael@epa.gov}\\

@@

@@

The standard GLM assumes the elements of \(\mathbf{y}\) are independent.
This independence assumption is typically impractical for spatial data.
For spatial data, nearby observations tend to be more similar than
distant observations \citep{tobler1970computer}, which leads to positive
spatial covariance among observations. The consequences of ignoring
spatial covariance in statistical models for spatial data can be severe
and include imprecise parameter estimates as well as misleading standard
errors that inflate Type-I error rates and decrease power
\citep{legendre1993spatial, zimmerman2024spatial}.

@@

Through inclusion of \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\), the spatial GLM (SPGLM) can be written as
\begin{align}\label{eq-spglm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
\end{align} Assuming independence between \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\), it follows that \begin{align}\label{eq-spcov}
 \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
\end{align} Henceforth, we refer to \(\sigma^2_{\tau}\) as
\(\sigma^2_{de}\) (for spatially dependent error variance) and
\(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for independent error
variance). The parameters \(\sigma^2_{de}\), \(\sigma^2_{ie}\), and
\(\phi\), in addition to any other parameters in \(\mathbf{R}\), compose
\(\boldsymbol{\theta}\), the covariance parameter vector.

@@

The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
recently released a full set of modeling tools for SPGLMs fit using the
novel Laplace approach described by \citet{ver2024marginal}. These
modeling tools are approachable and mirror the familiar \code{glm()}
syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs
relatively seamless. The \code{spglm()} function fits SPGLMs for
point-referenced support (e.g., x- and y-coordinates representing point
locations in a field; these models are sometimes called
``geostatistical'' models), while the \code{spgautor()} function fits
SPGLMs for areal support (e.g., polygon boundaries representing
geographic subsets of a region; these models are sometimes called
``autoregressive'' models). For both point-referenced and areal
supports, \pkg{spmodel} supports the binomial distribution for binary
data, Poisson and negative binomial distributions for count data, Gamma
and inverse Gaussian distributions for skewed data, and the beta
distribution for proportion data. There are 20 different spatial
covariance structures available including the exponential, Gaussian, and
spherical for point-referenced support (Figure\(~\)\ref{fig-type}) and
the conditional autoregressive and simultaneous autoregressive
structures for areal support. \pkg{spmodel} provides commonly used model
summaries, visualizations, and diagnostics (e.g., fitted values) using
standard \proglang{R} helper functions like \code{summary()},
\code{plot()}, and \code{fitted()}, among others. \pkg{spmodel} also has
tools to predict \(\mathbf{w}\) at new locations and quantify
uncertainty in those prediction using \code{predict()} and
\code{augment()}. This core functionality, combined with several
advanced features we describe throughout the manuscript, enables
\pkg{spmodel} to introduce novel, important SPGLM modeling tools
previously missing from the existing \proglang{R} ecosystem.

@@

Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version
0.11.1) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in
terms of scope and feel. Both packages use similar syntax to
\code{glm()}, accommodate flexible \code{formula} arguments (e.g.,
transformations, splines), handle spatial covariance that decays at
different rates in different directions (i.e., geometric anisotropy),
incorporate nonspatial random effects, support other \proglang{R}
packages for modeling like \pkg{broom}
\citep{robinson2021broom, kuhn2022tidy}, \pkg{emmeans}
\citep{lenth2024emmeans}, and \pkg{car} \citep{fox2019car}, and have
tools for model summaries, prediction, and simulating data. There are
some notable differences between the two packages, however. \pkg{sdmTMB}
supports several additional GLM distributions like the Tweedie, supports
Hurdle models, and can incorporate prior information through Bayesian
applications. \pkg{sdmTMB} also provides tools for working with temporal
data and spatiotemporal data and provides enhanced visualizations of the
model's marginal effects. \pkg{sdmTMB} does require a preprocessing step
of constructing a mesh prior to model fitting (using the stochastic
partial differential equation approach), and the density of the mesh can
affect model results and computational complexity. On the other hand,
\pkg{spmodel} does not require the construction of a mesh prior to model
fitting. \pkg{spmodel} supports 20 different spatial covariances and
models them directly, rather than using a precision matrix approximation
to the Matérn spatial covariance as in \pkg{sdmTMB}. \pkg{spmodel} can
model data using neighborhood distance and autoregressive models, rather
than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not
be within the polygon's boundaries. \pkg{spmodel} provides experimental
design tools (e.g., analysis of variance, contrasts), supports \pkg{sf}
objects in modeling and prediction functions \citep{pebesma2018sf}, has
several specialized model diagnostics like leverage values and Cook's
distances, and has analytic solutions for fixed effect and prediction
standard errors. Other similarities and differences do exist between
\pkg{sdmTMB} and \pkg{spmodel}, and both packages continue to evolve.
Overall, we believe that these packages are complementary and enhance
the suite of SPGLM tools accessible to practitioners.

@@

We can estimate the fixed effects using generalized least squares (GLS)
principles, a common practice for linear models estimated using
restricted maximum likelihood methods. Had we observed \(\mathbf{w}\), a
GLS estimator for \(\boldsymbol{\beta}\) is given by
\begin{align}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{align} where
\(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
While we do not know \(\mathbf{w}\), we do estimate it via
\(\hat{\mathbf{w}}\), and thus it is reasonable to define
\(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). To derive
properties of \(\hat{\boldsymbol{\beta}}\) like expectation and
variance, we must derive these properties for \(\hat{\mathbf{w}}\) by
conditioning on \(\mathbf{w}\) as if it were observed and leveraging the
laws of total expectation and variance. Based on asymptotic properties
of (restricted) maximum likelihood estimators
\citep{cressie1993asymptotic}, we may assume that given \(\mathbf{w}\),
\(\hat{\mathbf{w}}\) has mean \(\mathbf{w}\) and variance approximately
equal to \(-\mathbf{H}^{-1}\), the negative inverse Hessian (i.e., the
inverse observed information matrix). Thus it follows that
\(\text{E}(\hat{\mathbf{w}})\) is given by \begin{align}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{align} and \(\text{Var}(\hat{\mathbf{w}})\) is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & \approx \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align} Putting this all together, it follows that
\(\hat{\boldsymbol{\beta}}\) is unbiased for \(\boldsymbol{\beta}\):
\begin{align}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
\end{align} Moreover, it follows that \begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
\end{align} In practice, \(\text{Var}(\hat{\boldsymbol{\beta}})\) is
estimated by evaluating \(\boldsymbol{\Sigma}\) at
\(\hat{\boldsymbol{\theta}}\), the estimated covariance parameter
vector.

@@

Inspecting model diagnostics is an important step of the modeling
process that can yield valuable insights into model behavior and unusual
observations. \citet{montgomery2021introduction} contextualize three
components of unusual observations: outliers, leverage, and influence.
An observation is an outlier if it has an extreme response value
relative to expectation. The response GLM residuals simply compare the
observation to its fitted latent mean: \begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align} Because observations often have a unique support in a GLM
(e.g., only two possible response values for binary data) and the
variance of an observation generally depends on its mean, response
residuals lack some utility. Deviance, Pearson, and standardized
residuals are all different function of responses residuals that are
appropriately scaled to behave more like response residuals from a
standard linear model. For example, deviance residuals are given by
\begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align} where \(\mathbf{d}\) is a vector of individual deviances.
The sum of the squared deviance residuals equals the sum of the elements
of \(\mathbf{d}\), known as the deviance of the model fit. The deviance
of the model fit quantifies twice the difference in log likelihoods
between the a saturated model that fits every observation perfectly
(i.e., \(\text{y}_i = f^{-1}(\hat{\text{w}}_i)\) for all \(i\)) and the
fitted model \citep{myers2012generalized}. Deviance is often used as a
fit statistic, whereby lower values of deviance imply a better model fit
(given the observed data). Scaling the response residuals by the
reciprocal square root of the variance function (for the appropriate
response distribution) yields the Pearson residuals, and scaling the
Pearson residuals by the reciprocal square root of one minus the
leverage value (which we discuss next) yields the standardized residuals
\citep{faraway2016extending}.

An observation has high leverage if its combination of explanatory
variables is far away from other observations. In a linear model, the
leverage (i.e., hat) values are the diagonal of the leverage (i.e.,
projection, hat) matrix,
\(\mathbf{L} = \mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top\).
In a GLM, the leverage matrix is given by \begin{align}
  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2}.
\end{align} where \(\mathbf{V}\) is a diagonal matrix with \(i\)th
diagonal element equal to the variance function evaluated at
\(f^{-1}(\hat{\text{w}}_i)\). The larger the value of
\(\mathbf{L}_{ii}\), the more severe the leverage from the \(i\)th
observation.

@@

The \code{moose} data in \pkg{spmodel} contain information on moose
(Alces alces) presence in the Togiak region of Alaska, USA. \code{moose}
is an \code{sf} object, a special data frame that is supplemented with
spatial information using the \pkg{sf} package in \proglang{R}
\citep{pebesma2018sf}. After loading \pkg{spmodel}, the first few rows
of \code{moose} look like:

@@

Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel}
helper functions are available and that each of the \code{spglm()}
models uses the same hierarchical likelihood
(Equation\(~\)\ref{eq-marginal}):

@@

The \code{plot()} function can also be used to return similar
diagnostics as from \code{lm()} and \code{glm()}, with additional tools
for diagnosing spatial covariance. For example, we can inspect Cook's
distance values and the fitted spatial covariance as a function of
distance with (Figure\(~\)\ref{fig-sp-diagnostic2}):

@@

In addition to moose presence, moose counts are also recorded in
\code{moose} (Figure\(~\)\ref{fig-moose-data-count}). The Poisson and
negative binomial response distributions can be used to model SPGLMs for
count data. The Poisson distribution mean is equal to its variance,
while the negative binomial has an extra parameter to accommodate
overdispersion (where the variance is larger than the mean). Using a
spherical spatial covariance function, we may fit both a Poisson and
negative binomial SPGLM by changing the \code{family} argument:

@@

To model spatial dependence, we used a conditional autoregressive
function. Conditional and simultaneous autoregressive functions
characterize spatial distance through neighborhood relationships (rather
than Euclidean distance) and have \code{spcov_type} values of
\code{"car"} and \code{"sar"}, respectively. By default, Queen's
distance is used to determine whether two sites are neighbors, though
custom neighborhood matrices can be passed via the \code{W} argument.
Row standardization is also assumed by default; this can be changed via
the \code{row_st} argument. Using the \code{random} argument, we also
specified a nonspatial random effect for seal stock, which implies seals
belonging to the same stock share extra covariance. The \code{random}
argument uses similar syntax as \pkg{lme4} \citep{bates2015lme4} and
\pkg{nlme} \citep{pinheiro2006mixed} to specify nonspatial random
effects.

@@

According to AIC, the geostatistical SPGLM is preferred:

@@

The default estimation method in \pkg{spmodel} for SPGLMs is restricted
maximum likelihood (REML), while maximum likelihood (ML) can also be
used. A benefit of REML is that it can yield unbiased estimates of
covariance parameters \citep{cressie1993asymptotic}. A drawback of REML
is that likelihood-based statistics are often considered invalid when
the models have different explanatory variable or fixed effect
structures \citep{wolfinger1993covariance}, though
\citet{gurka2006selecting} provides some evidence to the contrary. In
contrast to REML estimators, ML estimators are generally biased for
covariance parameters, though in practice this bias tends to be small.
Moreover, when using ML, likelihood-based comparisons are valid for
models with different explanatory variable or fixed effect structures.
Using ML, we can evaluate the significance of log income on voter
turnout using a likelihood ratio test (for nested models):

\begin{CodeChunk}
\begin{CodeInput}
R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
R> spbeta_reduced_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
R> anova(spbeta_full_ml, spbeta_reduced_ml)
\end{CodeInput}
\begin{CodeOutput}
Likelihood Ratio Test
